---
title: "nb_background"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This experiment evaluates the FDP and power of different approaches to detecting
interaction effects in negative binomially distributed data. It is inspired by
the 3 vs 3 differential expression simulation setup from the clipper paper, but
with interaction effects between pairs of factors.

### Data Generation

We use the fact that the NB distribution arises from a poisson-gamma mixture.
The idea is to generate sample $i$ for gene $j$ from
\begin{align*}
y_{ji} \sim \text{NB}\left(\mu_{j}, \mu_{j}^{-1}\right)
\mu_{j}\left(x\right) &= X \beta_{j} \\
\beta_{j} &\sim \text{Gamma}\left(\alpha_{j}, \lambda_{j})
\end{align*}
where $X$ is a design matrix with all the main and interaction effects. If there
is no true interaction effect, then the corresponding coordinate of $\beta_{j} =
0$. Notice that their is variation even among the null genes -- this is similar
to the heterogeneous effect idea in the clipper simulations.

Here are functions we use to generate these data.
```{r}
set.seed(123)
library(tidyverse)
library(DESeq2)
```

```{r}
simulate_means <- function(X, assignment, hyper) {
  result <- matrix(nrow = length(assignment), ncol = nrow(X))
  for (j in seq_along(assignment)) {
    beta <- rgamma(ncol(X), hyper$alpha, hyper$beta)
    if (assignment[j] == "null") {
      beta[hyper$null_ix] <- 0
    } 
    result[j, ] <- X %*% beta
  }
  
  result
}

simulate_data <- function(mu, delta = 0.01) {
  N <- ncol(mu)
  J <- nrow(mu)
  y <- matrix(0, J, N)
  
  for (i in seq_len(N)) {
    for (j in seq_len(J)) {
      y[j, i] <- rpois(1, mu[j, i])
      if (runif(1) < delta) {
        y[j, i] <- 0
      }
    }
  }
  
  y
}
```

```{r}
n_samples <- 12
n_features <- 1e4
n_sig <- 1e3

hyper <- list(alpha = 25, beta = .5, null_ix = 4)
X <- matrix(1, nrow = n_samples, ncol = 4)
X[1:(n_samples / 2), 2] <- 0
X[, 3] <- c(rep(1, n_samples / 4), rep(0, n_samples / 4), rep(1, n_samples / 4), rep(0, n_samples / 4))
X[, 4] <- X[, 2] * X[, 3]
assignment <- c(rep("nonnull", n_sig), rep("null", n_features - n_sig))
coldata <- as.data.frame(X) %>%
  set_names("intercept", "factor1", "factor2", "interaction")

mu <- simulate_means(X, assignment, hyper)
y <- simulate_data(mu)
dds <- DESeqDataSetFromMatrix(
  countData = y,
  colData = coldata, 
  design = ~ 1 + factor1 + factor2 + factor1:factor2
) %>%
  DESeq()
res <- results(dds)
```
```{r}
output <- list()

n_reps <- 200
ix <- 1
qs <- seq(0.01, 0.2, by = 0.02)

for (rep in 1:n_reps) {
  for (delta in c(0, 0.01)) {
    print(rep)
    mu <- simulate_means(X, assignment, hyper)
    dds <- DESeqDataSetFromMatrix(
      countData = simulate_data(mu, delta = delta),
      colData = coldata, 
      design = ~ 1 + factor1 + factor2 + factor1:factor2
    ) %>%
      DESeq(fitType = "local")
    res <- results(dds)
    
    for (i in seq_along(qs)) {
      R <- which(res$padj < qs[i])
      output[[ix]] <- data.frame(
        "delta" = delta,
        "rep" = rep,
        "q"  = qs[i],
        "FDP" = length(intersect(R, (n_sig + 1):n_features)) / length(R)
      )
      ix <- ix + 1 
    }
  }
}

output <- bind_rows(output)
output$FDP[is.na(output$FDP)] <- 0 # happens when there are no rejections
```
```{r}
#write_csv(output, "~/Downloads/simulation_output.csv")
#output <- read_csv("~/Downloads/simulation_output.csv")
```


```{r, fig.width = 7, fig.height = 8}
min_theme <- theme_minimal() +
  theme(
    panel.background = element_rect(fill = "#ffffff"),
    panel.grid.major = element_line(color = "#e3e3e3"),
    panel.grid.minor = element_blank()
    )
theme_set(min_theme)
  
line_data <- data.frame(
  q = seq(0.01, .2, .02),
  FDP = seq(0.01, .2, .02)
)

ggplot(output) +
  geom_boxplot(aes(as.factor(q), FDP, fill = as.factor(delta)), size = 1, outlier.size = 2) +
  geom_line(data = line_data, aes(as.factor(q), FDP), group = 1, col = "#F24738", size = 1.5) +
  labs(
    x = "Target False Discovery Rate",
    y = "Actual False Discovery Proportion",
    fill = "Zero-Inflation Probability"
  ) +
  scale_fill_manual(values = c("#03588C", "#04C4D9")) +
  theme(
    axis.title = element_text(size = 15),
    axis.text = element_text(size = 12),
    legend.position = "bottom"
  )
ggsave("~/Downloads/interaction_fdr_experiment.png", dpi = 500, width = 5, height = 6)
  
```

