---
title: "nb_background"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This experiment evaluates the FDP and power of different approaches to detecting
interaction effects in negative binomially distributed data. It is inspired by
the 3 vs 3 differential expression simulation setup from the clipper paper, but
with interaction effects between pairs of factors.

### Data Generation

We use the fact that the NB distribution arises from a poisson-gamma mixture.
The idea is to generate sample $i$ for gene $j$ from
\begin{align*}
y_{ji} \sim \text{NB}\left(\mu_{j}, \mu_{j}^{-1}\right)
\mu_{j}\left(x\right) &= X \beta_{j} \\
\beta_{j} &\sim \text{Gamma}\left(\alpha_{j}, \lambda_{j})
\end{align*}
where $X$ is a design matrix with all the main and interaction effects. If there
is no true interaction effect, then the corresponding coordinate of $\beta_{j} =
0$. Notice that their is variation even among the null genes -- this is similar
to the heterogeneous effect idea in the clipper simulations.

Here are functions we use to generate these data.
```{r}
library(tidyverse)
signal_props <- list("1" = 10, "-1" = 10, "0" = 180)
```

```{r}
simulate_means <- function(X, signal_props, hyper) {
  signs <- map2(names(signal_props), signal_props, ~ rep(.x, .y)) %>%
    unlist() %>%
    as.numeric()
  
  D <- ncol(X)
  beta <- rgamma(D, hyper$alpha, hyper$beta)
  result <- matrix(nrow = length(signs), ncol = nrow(X))
  
  n_affected <- length(hyper$affected)
  for (j in seq_along(signs)) {
    if (signs[j] == 1) {
      beta[hyper$affected] <- rgamma(n_affected, hyper$up_scale * hyper$alpha, hyper$beta)
    } else if (signs[j] == -1) {
      beta[hyper$affected] <- rgamma(n_affected, hyper$down_scale * hyper$alpha, hyper$beta)
    } else {
      beta[hyper$affected] <- rep(0, n_affected)
    }
    
    result[j, ] <- X %*% beta
  }
  
  result
}

simulate_data <- function(mu) {
  N <- ncol(mu)
  J <- nrow(mu)
  y <- matrix(0, J, N)
  
  for (i in seq_len(N)) {
    for (j in seq_len(J)) {
      y[j, i] <- rnbinom(1, mu = mu[j, i], size = mu[j, i] ^ -1)
    }
  }
  
  y
}

n_samples <- 16
X_ <- data.frame(
  A = c(rep(-1, n_samples / 2), rep(1, n_samples / 2)),
  B = c(rep(-1, n_samples / 4), rep(1, n_samples / 4), rep(-1, n_samples / 4), rep(1, n_samples / 4))
)
X <- model.matrix(lm(rnorm(n_samples) ~ A * B, data = X_))
X[X == -1] <- 0

hyper <- list(alpha = 1, beta = 1, up_scale = 1.5, down_scale = 1.5 ^ -1, affected = 2:4)
mu <- simulate_means(X, signal_props, hyper)
y <- simulate_data(mu)
```

Genes 1 - 20 have real positive / negative effets. Everything else is a suprious
relationship.

```{r}
heatmap(y)
image(log(1 + t(y)))
```
